---
title: "Ch5-Homework"
author: "Isaac Kobby Anni"
date: "2024-10-10"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


<hr style=" height:4px; width:90%; border:3px soid red; border-radius:12px; background-color: black">

### Problem 2

```{r}
library(ggplot2)
library(dplyr)

```


**Data**
```{r}
compensation_amount <- c(11.73, 6.70660, 7.56180, 7.7738, 5.825, 7.121, 15.6062, 
                        12.0133, 10.7677, 7.0903, 4.3673, 14.3879, 6.5997, 6.38, 
                        6.0286, 10.7058, 17.558, 4.9357, 9.6019, 14.309, 9.3458, 
                        10.1015, 4.8093, 6.4407, 12.3189, 17.8454, 10.326, 2.2113, 
                        5.5036, 6.5355, 11.893, 3.109, 5.5714, 10.8262, 5.9958, 
                        4.2912, 2.0294, 8.0771, 5.3107, 9.252, 9.1823, 5.2043, 6.0456) # the values are divided by 100000
```

---

**a)** Produce a normal quantile plot or a normal probability plot. Add a normal distribution reference line to the plot.  
```{r}
qqnorm(compensation_amount,main=NULL,ylab = "Amount (Million US Dollars)",
       xlab="Standard Normal Quantiles",
       ylim=c(min(compensation_amount),max(compensation_amount)+3),xlim=c(-3,3))
#
abline(a=mean(compensation_amount),b=sd(compensation_amount), col='red', lwd=2.0)
```


---

**b)** Do the data in the plot of part (a) appear to be normally distributed? Discuss.     
- Yes, generally speaking, the plot indicates that reasonable amount of points lie on and/or are close to the reference line. Although there is slight skewness in the tails, it is not significant. The majority of the other points are nearly aligned with the reference line.

---

**c)** Would you choose to use your plot for part (a) in a written media presentation? If
not, then which type of plot would you use instead? Justify your choice.   
**Answer**:   
- No, because understanding this plot requires a solid background in statistics, which most of the media's audience, including the public, administrators, or those from other fields, may not have. Since these types of plots are rarely used for public presentations, a horizontal barplot or dotplot would be more suitable.  


<hr style=" height:4px; width:90%; border:3px soid red; border-radius:12px; background-color: black">


### Problem 4

**a)** Produce an EDF plot for the total compensation data. Add a normal curve to the EDF plot.

```{r}
#
mean_mass<-mean(compensation_amount)
sd_mass<-sd(compensation_amount)
min_mass<-min(compensation_amount)
max_mass<-max(compensation_amount)
#
xx<-seq(0,10000,1)*(max_mass-min_mass)/10000.+min_mass
yy<-pnorm(xx,mean_mass,sd_mass)

ecdfmass<-ecdf(compensation_amount)
plot.stepfun(ecdfmass,xlab="Amount (Million US Dollars)",
             ylab="Empirical Distribution Function",
             main=NULL,verticals=TRUE,do.points=FALSE,
             pch=16,xlim=c(0,max(compensation_amount)+3))

arrows(max(compensation_amount),1.0,1555,1.0,code=2,length=0.05)
arrows(-45.,0.0,min(compensation_amount),0.0,code=1,length=0.05)

```


**Adding a normal curve** 

```{r}
#
mean_mass<-mean(compensation_amount)
sd_mass<-sd(compensation_amount)
min_mass<-min(compensation_amount)
max_mass<-max(compensation_amount)
#
xx<-seq(0,10000,1)*(max_mass-min_mass)/10000.+min_mass
yy<-pnorm(xx,mean_mass,sd_mass)

ecdfmass<-ecdf(compensation_amount)
plot.stepfun(ecdfmass,xlab="Amount (Million US Dollars)",
             ylab="Empirical Distribution Function",
             main=NULL,verticals=TRUE,do.points=FALSE,
             pch=16,xlim=c(0,max(compensation_amount)+3))

arrows(max(compensation_amount),1.0,1555,1.0,code=2,length=0.05)
arrows(-45.,0.0,min(compensation_amount),0.0,code=1,length=0.05)

lines(xx,yy, col= "red")
```


**b)** Produce a normal quantile plot or a normal probability plot for the total compensation data. Add a normal distribution reference line to the plot. 

```{r}
qqnorm(compensation_amount,main=NULL,ylab = "Amount (Million US Dollars)",
       xlab="Standard Normal Quantiles",
       ylim=c(min(compensation_amount),max(compensation_amount)+3),xlim=c(-3,3))
#
abline(a=mean(compensation_amount),b=sd(compensation_amount), col='red', lwd=2)
```

**c)** Which of the two graphics from parts (a) and (b) would you prefer to represent to a class of students registered in a Master of Business Administration program? Discuss.    
 
- I would choose the Q-Q plot because it helps identify the type of distribution for a random variable, such as Gaussian, Uniform, Exponential, etc, by simply examining the plot. While MBA students are expected to understand both Q-Q plots and other types of visualizations, the second option is easier to interpret. It's important for MBA students to learn that clear and simple communication is key when presenting to their bosses and staff, so using easily interpretable plots is more favorable.

**d)** Is it reasonable to present either graphic from parts (a) or (b) to a public audience? Discuss.
- No this is not a good idea because misunderstandings can create a disconnect with the audience, so it's important to present our work as simple as possible. Both plots require a solid statistical background and are not suitable for a general or public audience. 



<hr style=" height:4px; width:90%; border:3px soid red; border-radius:12px; background-color: black">


### Problem 6  


```{r}
# Data
light <- c(28,32,31,34,21,26,40,25,39,22,30,25,30,30,
           27,19,27,23,36,27,24,25,33,16,37,24,26,24,
           25,26,29,31,23,16,28,33,28,26,27,29,32,29,
           28,21,36,25,29,36,29,20,26,36,27,-44,28,32,
           -2,28,24,32,32,23,22,28,24,27
           )
```

**a)** Produce an outlier boxplot for the data in Table 4.4. 

```{r}

out <- boxplot.stats(light)$out
boxplot(light,
  ylab = "Time",
  main = "",
  col = "lightblue",
  horizontal = TRUE
)
mtext(paste("Outliers: ", paste(out, collapse = ", ")))
```


 - We can see two outliers, -44 and -2 as shown in the diagram above.   

**b)** Produce a revised data set with the outliers removed
```{r}
Q1 <- quantile(light, 0.25)
Q3 <- quantile(light, 0.75)
IQR <- Q3 - Q1

lower_bound <- Q1 - 1.5 * IQR
upper_bound <- Q3 + 1.5 * IQR

light_no_outliers <- light[light >= lower_bound & light <= upper_bound]


cat('The revised data set: ', '\n' , light_no_outliers)
```
  

- As seen from the data, the -44 and -2 which were outliers are removed from the dataset. 

```{r}
out <- boxplot.stats(light_no_outliers)$out
boxplot(light_no_outliers,
  ylab = "Time",
  main = "",
  horizontal = TRUE
)
mtext(paste("Outliers: ", paste(out, collapse = ", ")))
```

- Again, the outlier boxplot shows no outliers exist in the revised version of the data. 

**c)** Produce a normal quantile plot, or a normal probability plot, for the data from your
answer to part (b). Add a normal distribution reference line to the plot.  
```{r}
qqnorm(light_no_outliers,main=NULL,ylab = "Amount (Million US Dollars)",
       xlab="Standard Normal Quantiles",
       ylim=c(min(light_no_outliers),max(light_no_outliers)+3),xlim=c(-3,3))
#
abline(a=mean(light_no_outliers),b=sd(light_no_outliers), col='blue', lwd=2)
```

**d)** Do the data in your plot for part (c) appear to be normally distributed? Discuss.

- Yes, the data appears to follow a normal distribution. Most of the points are closely aligned with the reference line, indicating normality. While a few points deviate slightly from the line, the overall distribution still suggests the data is normally distributed.




